---
title: "Violence in Mexico"
author: "Mariana Meza Hernandez"
date: "February 13, 2019"
output: 
  html_document:
    code_folding: hide
    css: style.css

---

```{r message= FALSE, warning = FALSE, echo=FALSE}
# load the required packages #
library(sf)
library(readr)
library(ggplot2)
library(tidyverse)
library(stringr)
library(extrafont)
library(ggpubr)
library(ggalluvial)
library(cowplot)
library(rnaturalearth)
library(rgeos)
library(scales)
library(waffle)
library(dplyr)
library(forcats)
library(knitr)
library(ggthemes)
```

```{r message= FALSE, warning = FALSE, echo=FALSE}
# generating my own theme
theme_mmh <- function(base_size = 16,
                      base_family = "Cambria",
                      base_line_size = base_size / 170,
                      base_rect_size = base_size / 170){
  theme_minimal(base_size = base_size, 
                base_family = base_family,
                base_line_size = base_line_size) %+replace%
    theme(
      
      plot.title = element_text(
        color = rgb(34, 34, 34, maxColorValue = 255), 
        face = "bold",
        hjust = 0,
        family = "Cambria"),
      plot.subtitle = element_text(
        family = "Garamond", 
        size = 12, 
        hjust = 0),
      plot.caption = element_text(
        family = "Garamond", 
        size = 10,
        hjust = 1),
      axis.title = element_text(
        color = rgb(73, 72, 72, maxColorValue = 255),
        size = rel(0.75),
        family = "Garamond"),
      axis.text = element_text(
        color = rgb(64, 64, 64, maxColorValue = 255),
        size = rel(0.65),
        family = "Garamond"),
      legend.title = element_text(
        color = rgb(64, 64, 64, maxColorValue = 255),
        size = rel(0.75),
        family = "Garamond"),
      legend.text = element_text(
        color =rgb(64, 64, 64, maxColorValue = 255),
        size = rel(0.75),
        family = "Garamond"),
      panel.background = element_blank(),
      panel.grid.major = element_line(
        rgb(192, 192, 187, maxColorValue = 255),
        linetype = "longdash"),   
     # panel.grid.minor = element_line(
      #  rgb(161, 160, 160, maxColorValue = 255),
       # linetype = "longdash", 
        #size = rel(4)),   
      
      complete = TRUE
    )
}
```

```{r, warning= FALSE, message= FALSE}
# load the datasets with direct paths
causegraph <- read_csv("~/Mariana Meza Hernandez/Harris/Winter 19/Data Visualiaztion/HW5/Data/causegraph.csv")
homicide <- read_csv("~/Mariana Meza Hernandez/Harris/Winter 19/Data Visualiaztion/HW5/Data/homicide_snsp_month.csv")
test <- read_csv("~/Mariana Meza Hernandez/Harris/Winter 19/Data Visualiaztion/HW5/Data/difftrendssnspinegi.csv")
rank_states <- read_csv("~/Mariana Meza Hernandez/Harris/Winter 19/Data Visualiaztion/HW5/Data/homicideraterankstate2.csv")
incidencia <- (read_csv("~/Mariana Meza Hernandez/Harris/Winter 19/Data Visualiaztion/HW5/Data/denunciados_incidencia.csv"))
insecurity <- read_csv("~/Mariana Meza Hernandez/Harris/Winter 19/Data Visualiaztion/HW5/Data/insecurityperception.csv")
mexstates <- (st_as_sf(ne_states(country = "mexico")))
sankey <- read_csv("~/Mariana Meza Hernandez/Harris/Winter 19/Data Visualiaztion/HW5/Data/sankey_graph_v2.csv")
mosaico <- list.files("~/Mariana Meza Hernandez/Harris/Winter 19/Data Visualiaztion/HW5/Data/mosaic", pattern = ".png", full.names = TRUE)
moran <- list.files("~/Mariana Meza Hernandez/Harris/Winter 19/Data Visualiaztion/HW5/Data/moran", pattern = ".png", full.names = TRUE)

```

## Homicide by firearm by month and year

Even though one cannot confirm causality between both events, it is clear that the level of homicides by firearm in Mexico increased dramatically since the start of the war against organized crime in early 2007. Over the last decades the number of homicides by firearm have remained almost four time higher than before the war began.

```{r, message=FALSE, warning = FALSE}
homicide_long <- gather (homicide, month, homicides, January:December, factor_key = TRUE) 
```

```{r warning= FALSE}
month_year_gun <-
  homicide_long %>%
  filter(subtype == "CON ARMA DE FUEGO") %>%
  filter (year > "2000") %>%
  group_by(year, month) %>%
  summarize(total_gun = sum(homicides))
```

```{r warning= FALSE, fig.width= 8, fig.height= 6}
mycols3 <- c("#7f1c32", "#2f2016", "#fcfaea", "#456789")
ggplot(data = month_year_gun) +
  geom_tile(aes(x = month, y =year, fill = total_gun)) +
  scale_y_continuous(breaks = 1997:2017, expand = c(0,0)) +
  scale_x_discrete(position = "top") +
  theme_mmh()+
  theme(axis.text.x = element_text(angle = 45, hjust = 0)) +
  geom_hline(aes(yintercept=2007), color="#222222") +
  scale_fill_gradient(low = "white", high = "#7f1c32") +
  ggtitle("Total homicides by firearm by month in Mexico") +
  labs(x = "Month", y = "Year", fill= "Homicides by firearm", caption = "Source: SNSP: Homicide records by the police", family= "Garamond")  +
   annotate("text", x = "May", y = 2006, label = "War Against Organized Crime Begins", size = 4.5, color = "#222222", family = "Garamond") 
```

## Difference between police reports and death certificates
The two main data sources that report homicides in Mexico are the homicide records by the police (SNSP) and the mortality statistics by the autonomous statistic institute (INEGI). In order to understand the violence in Mexico we need to look at both. One of the most interesting phenomena is that since the war against organized crime began, the trends in total violent homicides of both sources flipped. Since 2007, police reports have systematically undereported homicides in Mexico.

```{r, fig.width= 8, fig.height= 6}
ggplot(test, aes(x=Year, y = total, group=type))+
 geom_line(aes(color=type)) +
 geom_point(aes(color=type)) +
  theme_mmh()+
  theme(axis.text.x = element_text(angle = 45),
        panel.grid.major = element_blank(),
        legend.position = "none")+
  scale_color_manual(values=c("#7f1c32", "#2d7ac1"),
                     name = "Homicide records according to",
                     breaks = c("death_certificates", "police_reports"),
                     labels = c("Death certificates","Police reports"))+
  labs(x = "Year", y = "Total homicides" , subtitle = "This discrepancy is unusual and raises the question if local \n police and attorneys have deliberately underreported homicides", caption = "Sources: INEGI: Mortality statistics. \n and SNSP: Homicide records by the police") +

  xlab("Year") +
  ggtitle("Since 2007 total homicides reported by police are less \n  than the total number of homicide death certificates") +
  geom_vline(xintercept = 2007, linetype="dotted")+
  annotate("segment", x = 2008, xend = 2008, y = 13155, yend = 14006,
  colour = "#ff85a0", linetype= "dotted")+
  annotate("segment", x = 2009, xend = 2009, y = 16118, yend = 19803,
  colour = "#ff85a0", linetype= "dotted")+
  annotate("segment", x = 2010, xend = 2010, y = 20143, yend = 25757,
  colour = "#ff85a0", linetype= "dotted")+  
  annotate("segment", x = 2011, xend = 2011, y = 22409, yend = 27213,
  colour = "#ff85a0", linetype= "dotted")+ 
  annotate("segment", x = 2012, xend = 2012, y = 21459, yend = 25976,
  colour = "#ff85a0", linetype= "dotted")+ 
  annotate("segment", x = 2013, xend = 2013, y = 18106, yend = 23063,
  colour = "#ff85a0", linetype= "dotted")+ 
  annotate("segment", x = 2014, xend = 2014, y = 15520, yend = 20010,
  colour = "#ff85a0", linetype= "dotted")+ 
  annotate("segment", x = 2015, xend = 2015, y = 16909, yend = 20762,
  colour = "#ff85a0", linetype= "dotted")+ 
  annotate("segment", x = 2016, xend = 2016, y = 20547, yend = 24559,
  colour = "#ff85a0", linetype= "dotted")+ 
  annotate("segment", x = 2017, xend = 2017, y = 25339, yend = 32079,
  colour = "#ff85a0", linetype= "dotted")+ 
  annotate("text", x = 2007, y = 25000, label = "War Against Organized Crime", angle=90, family = "Garamond", color = "darkgrey" ) +
  annotate("text", x = 2000, y = 17000, label = "Police reports",  family = "Garamond", color = "dodgerblue4" ) +
  annotate("text", x = 2011, y = 29500, label = "Death certificates",  family = "Garamond", color = "darkred" ) +
  scale_x_continuous(breaks = 1997:2017, expand = c(0,0.5))
 
```

## Homicide rates by state through the years
The dynamics of homicides have a very local component. As the graph shows, homicides have increased in some states like Colima, Baja California Sur and Zacatecas while it has decreased in states like Durango, the State of Mexico or Queretaro. This is in part because of the varying strength and violence of the drug cartels, and the direct confrontations between the cartels and the Mexican police and army.

```{r message=FALSE, fig.width= 10, fig.height= 8}
colors <- c("#DBDBDB","#DBDBDB","#7f1c32","#DBDBDB","#2d7ac1","#DBDBDB","#DBDBDB","#ff3864","#3ba1ff","#DBDBDB","#DBDBDB","#DBDBDB","#DBDBDB","#87c6ff","#DBDBDB","#DBDBDB","#DBDBDB","#DBDBDB","#DBDBDB","#DBDBDB","#DBDBDB","#DBDBDB","#DBDBDB","#DBDBDB","#DBDBDB","#DBDBDB","#DBDBDB","#DBDBDB","#DBDBDB","#DBDBDB","#DBDBDB","#ff85a0")
ggplot(data = rank_states, aes(x = year, y = rank, group = states)) +
  geom_line(aes(color = states, alpha = 1), size = 2) +
  geom_point(aes(color = states, alpha = 1), size = 1) +
  scale_y_reverse(breaks = 1:nrow(rank_states), expand = c(0,0.5))+
  geom_point(color = "#FFFFFF", size = 1)+
  scale_x_continuous(breaks = c(2007,2009,2012,2015,2017), expand = c(0.1,0.05))+
  geom_text(data=rank_states %>% filter(year=="2007"),
            aes(label=states, x=2006), size=3.5, family = "Garamond", color = "#222222")+
  geom_text(data = rank_states %>% filter(year == "2017"),
            aes(label = states, x = 2018) , size = 3.5, family = "Garamond", color = "#222222") +
  coord_cartesian(ylim = c(1,32)) + 
  theme_mmh() +
  theme(legend.position = "none",
        axis.title.y=element_blank(),
        plot.title = element_text(hjust=0),
        plot.subtitle = element_text(hjust=0),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_color_manual(values = colors)+
  labs(x = "Year",
       title = "Relocation of the Zetas Cartel from Tamaulipas to Colima and Zacatecas caused \n an increase in homicides",
       subtitle = "The states in red colors show the bigest relative increase in homicide rate.\nStates in blue colors show the biggest relative decrease.", caption = "Source: SNSP: Homicide records by the police", family = "Garamond")
``` 

## Municipalities by level of violence
The municipalities with the largest homicide rates per 100,000 are located in the West and Southeast regions of the country. The Northwest region comes in a  third place. The graph shows how homicides are concentrated in particular very violent municipalities in each region. A targeted security policy in those municipalities is key.

```{r out.width = "150%"}
knitr:: include_graphics(mosaico)
```

## Spatial autocorrelation of homicide levels between municipalities
One of the most important measures of spatial correlation is Local Moran's I. 
Local Moran's I measures how geographical units can form statistically significant clusters. A positive value for Moran's I indicates that a feature has neighboring features with similarly high or low attribute values; this feature is part of a cluster. In this case the dark red or dark blue clusters are clusters with similar neighbors. A negative value for I indicates that a feature has neighboring features with dissimilar values; this feature is an outlier. For the case of the municipalities, the light red and light blue are clusters with dissimilar neighbors.

```{r out.width = "150%"}
include_graphics(moran)
```

## Perception of insecurity in states

When people are asked to describe the level of insecurity of their state, their perception varies significantly from state to state.
```{r message= FALSE}
mexstates <- left_join(mexstates, insecurity)
mexstates[12,86] = 3
```

```{r fig.width= 8, fig.height= 6}
ggplot(mexstates, aes(fill=factor(rank_perc))) +
  geom_sf() +
  scale_fill_manual(breaks = c(1,2,3,4),
                    labels = c("Low", "Medium", "High","Very High"),
                        values=c("white", "#ff85a0", "#ff3864", "#7f1c32")) +
  labs(title="Perception of insecurity varies within states",
       fill = "Perception of insecurity in your state?",
       caption = "Sources: ENVIPE:Perception of insecurity.",
       subtitle = "More people in Veracruz and Mexico City consider their state \n to be insecure than in Yucatan and Campeche")+
  theme_mmh()+
  theme(axis.title.x=element_blank(),
        plot.title = element_text(hjust = 0),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "top")
```

## Cause of homicide by sex

The way men and women are killed show different patterns that may point to different phenomena. Women, unlike men, are killed by asphyxiation and knife that point towards the high levels of feminicides in the country. Men on the other hand are killed mostly by firearm that point towards the deaths involved in the war of drugs. 
```{r warning = FALSE, message= FALSE}
ggplot(causegraph) +
geom_bar(mapping = aes(x = reorder(cause3, -percent), y = percent, group = sexo2, fill = factor(sexo2)),stat = "identity", position = "dodge")+ coord_flip() +
  scale_y_continuous(expand = c(0,1))+
  scale_fill_manual ("Gender", values = c("Men" = "#2d7ac1", "Women" = "#7f1c32")) +
  labs(x = "Cause of homicide", y = "Percent", subtitle = "Men and women are killed primarely by firearm. Yet women are more likely \n than  men to be murdered by asphyxiation or knife, indicating potential feminicides", caption = "Source: INEGI: Mortality Statistics") +
  ggtitle("Women are more likely to be  killed by knife \n or asphyxiation than men") +
  guides(fill=guide_legend("Gender")) +
  theme_mmh()
```

## Place of homicide and relationship to killer by gender

Another difference between men and women is that most women are killed at home, regardless if they are killed by a relative or not. Furthermore, more women are killed by relatives than strangers. In contrast, men are killed predominantly in the streets either by relatives or strangers.

```{r}
colors <- c("#2d7ac1", "#ff3864", "#ccc9b3")
ggplot(sankey,
       aes(y = freq,
          axis1 = location, axis2 = related, axis3 = sexo)) +
  geom_alluvium(aes(fill = sexo),
                width = 0, knot.pos = 0, reverse = FALSE) +
  guides(fill = FALSE) +
  geom_stratum(width = 1/8, reverse = FALSE) +
  geom_text(stat = "stratum", label.strata = TRUE, reverse = FALSE, family = "Garamond") +
  scale_x_continuous(breaks = 1:3, labels = c("Location", "Related", "Sex")) +
  coord_flip() +
  ggtitle("  Women are more likley to be killed by a relative at home
          Men are more likely to be killed in the street")+
  theme_mmh()+
  theme(
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
  ) +
  labs(caption = "Sources: INEGI: Mortality statistics.") +
  scale_fill_manual(values = colors) 

```